Neighbor-Based Decentralized Training Strategies for Multi-Agent Reinforcement Learning

### 2024-10-12 paper-2024-sac-marl-neighbouring-strategies.tex
pass till 2.3

\begin{abstract}
Multi-agent deep reinforcement learning (MADRL) is rapidly emerging as a powerful extension of single-agent reinforcement learning, 
enabling the training of both collaborative and competitive agents in several domains, like robotics, traffic management, and video games.
%
The most widely adopted training paradigm in MADRL is centralized training with decentralized execution (CTDE), which has demonstrated significant effectiveness. 
%
However, it relies on a centralized learner, requiring agents to learn in an offline setting and execute online. To address this limitation, decentralized training methods have been introduced. 
%
While these techniques offer greater flexibility, they tend to have slower convergence and can suffer from instability compared to centralized approaches.
%
This paper investigates the effectiveness of various neighbor-based decentralized training strategies based 
  on the well-known Deep-Q Learning algorithm as a viable alternative to centralized training. 
%  
We evaluate experience sharing, k-nearest neighboring averaging, and k-nearest neighboring consensus methods 
  in a custom multi-agent environment and compare their performance against centralized training
  and totally decentralized training. 
%  
Our results show that neighbor-based methods can achieve comparable performance to centralized training 
  while offering improved scalability and communication efficiency. 
%
% Finally, we discuss the trade-offs between these methods and provide insights into their applicability 
%   in different scenarios.
\end{abstract}

@@

\section{Background and Motivation}\label{sec:background}
MARL is an extension of a sequential decision-making process, 
where multiple \emph{agents} interact with each other and a shared \emph{environment}.
In particular, each agent in the system aims to maximize its own \emph{reward signal} (i.e., a scalar feedback signal),
which is typically a function of the global state (sometime not known, making the system \emph{partial observable}) and the joint action taken by all agents.
%
In doing that, each agent perceive the environment through its own \emph{observation function},
which maps the global state to the agent's local observation and perform an action based on its \emph{policy} (i.e., a mapping from observations to actions).
%
This chosen action is then combined with the actions of other agents forming a \emph{joint action},
which is used to update the global state and reward.

@@

\begin{itemize}
  \item $\mathcal{G} = (N, E)$ is a communication graph, where $N$ is the set of $n$ agents and $E \subseteq N \times N$ represents the communication links between agents. 
  Time-varying graphs $\mathcal{G}_t = (N, E_t)$ can be used to represent communication evolving over time $t$.
  \item $\mathcal{S}$ is the global state space.
  \item $\mathcal{A} = \mathcal{A}^1 \times \dots \times \mathcal{A}^n$ is the joint action space, where $\mathcal{A}^i$ is the action space of agent $i$.
  \item $\mathcal{O} = \mathcal{O}^1 \times \dots \times \mathcal{O}^n$ is the joint observation space, where $\mathcal{O}^i$ is the observation space for agent $i$.
  \item $\mathcal{P}: \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is the state transition function, describing the probability of transitioning to a new state $s' \in \mathcal{S}$ given the current state $s \in \mathcal{S}$ and joint action $a \in \mathcal{A}$.
  \item $\mathcal{R} = {\mathcal{R}^i}_{i \in N}$, where $\mathcal{R}^i: \mathcal{S} \times \mathcal{A} \to \mathbb{R}$ is the reward function for agent $i$.
  \item $\gamma \in [0, 1]$ is the discount factor.
\end{itemize}
Each agent $i$ at time $t$ receives an observation $o^i_t \in \mathcal{O}^i$, 
takes an action $a^i_t \in \mathcal{A}^i$ based on its local policy $\pi^i: \mathcal{O}^i \times \mathcal{A}^i \to [0,1]$, and receives a reward $r^i_{t+1} = \mathcal{R}^i(s_t, a_t)$. 
%
The global state evolves according to $\mathcal{P}$. 
Agents can exchange information with their neighbors in $\mathcal{G}$ 
(or $\mathcal{G}t$ if the graph is time-varying). 
%
The objective is typically to maximize the expected discounted sum of rewards $\sum{t=0}^{\infty} \gamma^t r^i_t$ for each agent $i$, 
or some global reward function like the average reward over all agents.
Formally, it can be defined as:
\[ J(\{\pi^i\}) = \lim_{T \to \infty} \mathbb{E} \left[ \frac{1}{T} \sum_{t=0}^{T-1} \frac{1}{N} \sum_{i \in \mathcal{N}} R^i(s_t, a_t) \right] \]

In doing this, agents typically use other function to estimate the value of taking an action in a given state, such as the action-value function (also known as Q-value). 
The latter represents the expected cumulative discounted reward the agent will receive by taking action $a \in \mathcal{A}$ in state $s \in \mathcal{S}$ and then following its policy $\pi^i$:

@@

%where $\pi = \{\pi^i\}_{i \in N}$ represents the joint policy. 
%The goal in collaborative MARL is often to find a joint policy that maximizes this global average Q-value or a similar global objective.
Several algorithms aim to find the \emph{optimal} Q-value function, which maximizes the expected cumulative reward. One of the most popular among these is the Deep Q-learning algorithm.
\subsubsection{Deep Q-Learning Formalization}
Deep Q-learning~\cite{DBLP:conf/l4dc/FanWXY20} utilizes a deep neural network, 
often referred to as a Q-network, to approximate the action-value function $Q(s, a)$. 
This network takes the state $s$ as input and outputs a vector of Q-values, one for each possible action $a \in \mathcal{A}$.  
The optimal action-value function, denoted $Q^*(s, a)$, satisfies the Bellman optimality equation:

@@

\paragraph{Centralized Training with Decentralized Execution (CTDE)}
In CTDE, a central learner trains all agents, 
but execution of the learned policy is decentralized. 
This process typically involves two phases:
\begin{description}
  \item[Offline training:] Agents interact with a simulator or environment, gathering experience data used to update the central learner's policy.
  \item[Online execution:] Agents deploy the learned policy in the environment without further learning.
\end{description}
This approach offers several benefits.  
Leveraging a centralized dataset allows for efficient learning, 
potentially exploiting global information, 
which is crucial in cooperative scenarios requiring coordinated actions. 
Furthermore, decentralized execution allows agents to act independently after training.
%
However, CTDE has limitations. 
Scalability can become problematic with a large number of agents. 
The reliance on a central learner hinders adaptation to dynamic environments or tasks during runtime.
\sloppy
\paragraph{Decentralized Training with Decentralized Execution (DTDE)}
DTDE eliminates the central learner; each agent learns independently from its own experiences.
Inter-agent cooperation and communication can be incorporated during both training and execution. 
Unlike CTDE, training and execution phases are not necessarily separate---
simultaneous learning and execution allow for adaptation to changing environments or tasks at runtime.
%
Key advantages of DTDE include scalability with the number of agents and adaptability to dynamic environments.
%
However, DTDE faces challenges. 
Convergence time is often longer compared to centralized training. 
The lack of central coordination can lead to instability and difficulties in differentiating agents and their environmental interactions. 
Online learning in DTDE can introduce non-stationarity if agents do not account for other agents' evolving policies during training, 
as the environment effectively changes from each agent's perspective.

>>>>

\begin{abstract}
Multi-agent deep reinforcement learning (MADRL) is rapidly emerging as a powerful paradigm for training intelligent agents in complex, multi-agent environments.
%
Extending beyond single-agent reinforcement learning, MADRL enables the development  of both collaborative and competitive agents in several domains, like robotics, traffic management, and video games.
%
The most widely adopted training paradigm in MADRL is centralized training with decentralized execution (CTDE), which has demonstrated significant effectiveness. 
%
However, its reliance on a centralized learner necessitates that agents acquire policies offline and subsequently execute them online. 
This inherent constraint motivates the exploration of decentralized training methodologies.
%
While these techniques offer greater flexibility, they tend to have slower convergence and can suffer from instability compared to centralized approaches.
%
Therefore, this paper investigates the effectiveness of various neighbor-based decentralized training strategies based 
  on the well-known Deep-Q Learning algorithm as a viable alternative to centralized training. 
%  
We evaluate experience sharing, k-nearest neighboring averaging, and k-nearest neighboring consensus methods 
  in a cooperative multi-agent environment and compare their performance against centralized training
  and totally decentralized training. 
%  
Our results show that neighbor-based methods can achieve comparable performance to centralized training 
  while offering improved scalability and communication efficiency. 
%
% Finally, we discuss the trade-offs between these methods and provide insights into their applicability 
%   in different scenarios.
\end{abstract}

@@

\section{Background and Motivation}\label{sec:background}
Multi-agent reinforcement learning (MARL) involves a group of \emph{agents} interacting with a shared \emph{environment}. 
%
At each timestep, agents receive \emph{observations} and select \emph{actions} according to their individual \emph{policies} (mappings from observations to actions). 
%
These individual actions combine to form a \emph{joint action}, which influences the environment's transition to a new state. 
%
Each agent then receives a scalar \emph{reward} signal, typically a function of the global state and joint action. 
%
The goal of each agent is to maximize its cumulative reward. 
%
Environments can be \emph{partially observable}, meaning agents may not have access to the full global state, only receiving local observations through their respective \emph{observation functions}.

@@

\begin{itemize}
  \item $\mathcal{G} = (N, E)$ is a communication graph, where $N$ is the set of $n$ agents and $E \subseteq N \times N$ represents the communication links between agents. 
  Time-varying graphs $\mathcal{G}_t = (N, E_t)$ can be used to represent communication evolving over time $t$.
  \item $\mathcal{S}$ is the global state space.
  \item $\mathcal{A} = \mathcal{A}^1 \times \dots \times \mathcal{A}^n$ is the joint action space, where $\mathcal{A}^i$ is the action space of agent $i$.
  \item $\mathcal{O} = \mathcal{O}^1 \times \dots \times \mathcal{O}^n$ is the joint observation space, where $\mathcal{O}^i$ is the observation space for agent $i$.
  \item $\mathcal{P}: \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is the state transition function, describing the probability of transitioning to a new state $s' \in \mathcal{S}$ given the current state $s \in \mathcal{S}$ and joint action $a \in \mathcal{A}$.
  \item $\mathcal{R} = {\mathcal{R}^i}_{i \in N}$, where $\mathcal{R}^i: \mathcal{S} \times \mathcal{A} \to \mathbb{R}$ is the reward function for agent $i$.
  \item $\gamma \in [0, 1]$ is the discount factor.
\end{itemize}
Each agent $i$ at time $t$ receives an observation $o^i_t \in \mathcal{O}^i$, 
takes an action $a^i_t \in \mathcal{A}^i$ based on its local policy $\pi^i: \mathcal{O}^i \times \mathcal{A}^i \to [0,1]$, and receives a reward $r^i_{t+1} = \mathcal{R}^i(s_t, a_t)$. 
%
The global state evolves according to $\mathcal{P}$. 
Agents can exchange information with their neighbors in $\mathcal{G}$ 
(or $\mathcal{G}t$ if the graph is time-varying). 
%
The objective is typically to maximize the expected discounted sum of rewards $\sum{t=0}^{\infty} \gamma^t r^i_t$ for each agent $i$, 
or some global reward function like the average reward over all agents.
Formally, it can be defined as:
\[ J(\{\pi^i\}) = \lim_{T \to \infty} \mathbb{E} \left[ \frac{1}{T} \sum_{t=0}^{T-1} \frac{1}{N} \sum_{i \in \mathcal{N}} R^i(s_t, a_t) \right] \]
Agents typically use other function to estimate the value of taking an action in a given state, such as the action-value function (also known as \emph{Q-value}). 
The latter represents the expected cumulative discounted reward the agent will receive by taking action $a \in \mathcal{A}$ in state $s \in \mathcal{S}$ and then following its policy $\pi^i$:

@@

%where $\pi = \{\pi^i\}_{i \in N}$ represents the joint policy. 
%The goal in collaborative MARL is often to find a joint policy that maximizes this global average Q-value or a similar global objective.
Several algorithms aim to find the \emph{optimal} Q-value function, which maximizes the expected cumulative reward. 
%
One of the most popular among these is the Deep Q-learning algorithm.
\subsubsection{Deep Q-Learning Formalization}
Deep Q-learning~\cite{DBLP:conf/l4dc/FanWXY20} utilizes a deep neural network, 
often referred to as a Q-network, to approximate the action-value function $Q(s, a)$. 
This network takes the state $s$ as input and outputs a vector of Q-values, one for each possible action $a \in \mathcal{A}$.  
The optimal action-value function, denoted $Q^*(s, a)$, satisfies the Bellman optimality equation:

@@

\paragraph{Centralized Training with Decentralized Execution (CTDE)}
In CTDE, a central learner trains all agents, 
but execution of the learned policy is decentralized. 
This process typically involves two phases:
\begin{itemize}
  \item \textit{Offline training:} Agents interact with a simulator or environment, gathering experience data used to update the central learner's policy.
  \item \textit{Online execution:} Agents deploy the learned policy in the environment without further learning.
\end{itemize}
This approach offers several benefits.  
Leveraging a centralized dataset allows for efficient learning, 
potentially exploiting global information, 
which is crucial in cooperative scenarios requiring coordinated actions. 
Furthermore, decentralized execution allows agents to act independently after training.
%
However, CTDE has limitations. 
Scalability can become problematic with a large number of agents. 
The reliance on a central learner hinders adaptation to dynamic environments or tasks during runtime.
\sloppy
\paragraph{Decentralized Training with Decentralized Execution (DTDE)}
DTDE eliminates the central learner; each agent learns independently from its own experiences.
Inter-agent cooperation and communication can be incorporated during both training and execution. 
Unlike CTDE, training and execution phases are not necessarily separate---simultaneous learning and execution allow for adaptation to 
changing environments or tasks at runtime.
%
Key advantages of DTDE include scalability with the number of agents and adaptability to dynamic environments.
%
However, DTDE also presents several challenges to overcome, like: 
i) convergence time is often longer compared to centralized training. 
ii) The lack of central coordination can lead to instability and difficulties in differentiating agents and their environmental interactions; and 
iii) online learning in DTDE can introduce non-stationarity if agents do not account for other agents' evolving policies during training, 
as the environment effectively changes from each agent's perspective.
